---
title: "Interactive Visualization of US Monthly Birth Data (2000-2014) along with predection scatter plot & Lakes in Florida - Spatial Visulization"
author: "Divyesh Desai - `ddesai7656@floridapoly.edu`"
output: html_notebook
---
***

> Importing the required library.

```{r}
library(tidyverse)
library(plotly)
library(sf) # For spatial data
library(ggplot2)
```

> Loading the data from the data directory and reading it through the imported library.


```{r}
birth_data_url <- "../data/us_births_00_14.csv"
births <- read.csv(birth_data_url, stringsAsFactors = FALSE)
```

> filtering data to create plot

```{r}
monthly_births <- births %>%
  group_by(year, month) %>%
  summarize(total_births = sum(births), .groups = 'drop') %>%
    mutate(date = as.Date(paste(year, month, "01", sep = "-")),
         month_name = format(date, "%B"))
```

### Visualization 2 - Linear Model: Total Births Over Years


> generating interactive plot from the filtered data

```{r}
plot <- plot_ly(monthly_births, x = ~year, y = ~total_births, type = 'scatter', mode = 'lines+markers', text = ~paste("Year:", year, "<br>Month:", month_name, "<br>Total Births:", total_births), 
                hoverinfo = 'text') %>%
  layout(
    title = "Total Monthly Births in the US (2000-2014)",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Total Births")
  )
plot
```

### Visualization 2 - Linear model: Predicting total births based on year


> Linear model: Predicting total births based on year


```{r}
model <- lm(total_births ~ year, data = monthly_births)
```


> Summary of the model

```{r}
summary(model)
```


> Plot the model with ggplot2


```{r}
ggplot(monthly_births, aes(x = year, y = total_births)) +
  geom_point() +
  geom_smooth(method = "lm", col = "#532D8E") +
  labs(title = "Linear Model: Total Births Over Years", x = "Year", y = "Total Births") +
  theme_minimal()
```


### Visualization 3 - Lakes in Florida


> Reading the shape file from the data directory

```{r}
lakes <- st_read("../data/Florida_Lakes/Florida_Lakes.shp")
```

> mapping lakes to the florida map using red color

```{r}
ggplot() +
  geom_sf(data = lakes, fill = "red", color = "black") +
  ggtitle("Lakes in Florida") +
  theme_minimal()
```